---
title       : "NMR multiplet simulator"
author      : "Paolo Pirovano"
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

<style>
.small-code pre code {
  font-size: 1em;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=20, fig.height=4)
load("../figures.Rdata")
source("../scripts/plotters.R")
source("../scripts/couplers.R")
source("../scripts/lbroad.R")
source("../scripts/nmrsim1.R")
library(ggplot2)
```

## What is this app about?


- A Shiny app to simulate the fine structure in NMR spectra
- This has got nothing to do with statistics or data science
- I had it half-done already from a personal project

### So, um, what's an NMR?


- No need to worry, just take it as a squiggly line generator
- https://en.wikipedia.org/wiki/Nuclear_magnetic_resonance_spectroscopy
- https://en.wikipedia.org/wiki/J-coupling

Why would I want it?

- Chemistry/physics homework maybe?

--- .small-code
## How do I use it?

Input your list of nuclides in the table

- I = nuclear magnetic moment (0.5, 1, 1.5, ...)
- J = Coupling constant (whatever you want)
- n = number of equivalent nuclides (positive integer)

``` {r}
simspec <- function(nuclides, lb = 1) {
      spectrum <- data.frame(x = 0, y = 1)
      for (i in 1:nrow(nuclides)) {
            spectrum <-
                  couplen(spectrum,
                          I = nuclides$I[i],
                          J = nuclides$J[i],
                          n = nuclides$n[i])
      }
}
```

---

## Example simulation

``` {r fig.width=10, fig.height=3}
nuclides <-  data.frame(I = c(.5, 1), J = c(7, 20), n = c(3,1))
spectrum <- simspec(nuclides)
plotSpec(spectrum[[1]])
```

---

## Line broadening

We get finite linewidth with a gaussian convolution (?) of the spectral lines.
You can select the broadening with the slider.

``` {r, echo = FALSE, fig.width=10, fig.height=3}
plotSpec(figures[[1]])
plotSpecLb(figures[[2]])
```



